Darwin's Dream: Unifying ecological and evolutionary change

### 2025-05-29 response_to_reviewers.md

We thank both reviewers for their valuable comments and suggestions. They helped us to improve the presentation of a number of points that were not entirely clear in the previous version of the manuscript and focus our revision on better communicating the aim of our work. To summarise, it does not appear that the accuracy or novelty of our proposed equation is in question. But the motivation for seeking and reporting it clearly needs to be better communicated. Couching this motivation in terms of major historical breakthroughs in science has set an incorrect and unintended tone; the point was to explain why pursuing abstract unifying equations is a valid and important part of scientific theory (using the strongest cases available), not to suggest that what we've done is field-changing in its importance!

Consequently, the references to major advances in science, which we appreciate came across as self-aggrandising, were motivated primarily by what we feel is a need to justify the importance and validity of a particular kind of theory. We have scaled back on this emphasis, but we feel as though some such justification is still necessary given the nature of comments from both reviewers, with Reviewer 1 closing their review with "[e]xplaining why it might be useful would help too", and Reviewer 2 stating that "if this equation does not produce evolutionary or ecologically meaningful results on its own, it's unclear that it has any utility". While we believe that there is scope to derive pragmatic results from our equation (i.e., new predictions), the primary goal of this manuscript is, unapologetically, about conceptual and theoretical clarity rather than prediction. This is not different from George Price's works from the 1970s. For example, in his 1972 paper he states that his now famous formula "is intended mainly for use in deriving general relations and constructing theories, and to clarify understanding of selection phenomena, rather than for numerical calculation" (Price 1972, 485). In this sense, our manuscript follows a similar path. We ask, what is most fundamental to all eco-evolutionary change? And we deliver a formal answer, the accuracy of which does not appear to be in question. Ideally, all formalisms try to achieve three characteristics at the same: generality, realism, and precision (Levins 1966). However, as Levins (1966) points out, it is not possible to achieve all of them at the same time and one of them needs to be sacrificed. When you theorise the way that we do here, specific (precise) predictions are sacrificed in favour of generality and realism (more on this later in relation to "dynamic sufficiency").

We contend that this kind of theory is important in ecology and evolution, even if it is not of widespread interest or appreciation (Caswell 1988). For example, even if the Price equation served no other purpose than to provide a formal, abstract, link between the fundamental equations of population and quantitative genetics (Queller 2017), it would still be a valid contribution to evolutionary biology, useful for providing conceptual clarity to theoreticians and demonstrating the logical consistency of the field. Hence, while we are optimistic about where our starting point might lead for developing new theory, we do not see this as being the most important or interesting outcome of this work.

Our equation is general and realistic. It is also simple; it couldn't be anything else given what we set out to achieve (and while it is perhaps trivial to verify in hindsight, it was not trivial to find). A fundamental equation of eco-evolutionary change must, by definition, be able to recover the fundamental equations of evolutionary biology (the Price equation) and population ecology (N_{t+1} = N_{t} + Births - Deaths). Our equation thereby defines what eco-evolutionary change is in its broadest sense; verbally, it is an interaction between individual fitness components (birth, death, and identity) and an individual characteristic.

Reviewer 1 comment 1
==================================

The literature that is cited is VERY selective. There is a huge body of literature on linking ecological and evolutionary theory. High-dimensional matrix models have now been routinely used to link ecological and evolutionary dynamics by Barfield, Boots, Childs, Coulson, Ellner, Lion, Rees, Tuljapurkar and others. They have shown how by writing equations to dynamically track distributions of individual-level phenotypic traits and genotypes you can retrieve: the Price equation (including for structured populations) selection differentials and gradients, heritabilities, genetic covariances, the Euler-Lotka equation, equations for population growth, rates of evolutionary change, evolutionary end points, species interactions, and coexistence mechanisms. This literature needs citing and appropriately summarising. (It turns out it is straightforward to retrieve their equation from this modelling framework. if the authors did this, it will help reveal its limitations. One example is their delta zi terms get split into ontogenetic development among surviving individuals and genetic and non-genetic inheritance between parents and their offspring in age-structured populations).

Author response to Reviewer 1 comment 1
=====================================

We have revised our introduction extensively to address this comment and better communicate the goals of our work.

We agree with Reviewer 1 concerning the existence and importance of this literature integrating ecological and evolutionary theory. We are familiar with this literature, but all of the models developed by the authors mentioned by Reviewer 1, to our knowledge, are addressing questions that are fundamentally different from our own. These are highly general tools for integrating ecological and evolutionary dynamics, and we cited some key ones in our original introduction (Lion 2018; Patel, Cortez, and Schrieber 2018; Govaert et al. 2019). Lion (2018) and Patel et al. (2018) are particularly instructive examples because of how general they are: In Lion (2018), equations for ecological [ dn_{i}/dt = r_{i}(E)n_{i} ] and evolutionary [ f_{i}(r_{i}(E) - \bar{r_{i}(E)) ] change are defined as functions of an environmental vector E that can include other population densities or any relevant external variables. In Patel et al. (2018), ecological [ dN_{i}/dt = N_{i} f_{i}(N, x) ] and evolutionary [ dx_{j}/dt = e g_{j}(N, x) ] equations are similarly defined by general functions for population change (f) and trait change (g). Govaert et al. (2019) reviews eco-evolutionary feedback models.

There are, of course, other excellent examples by the authors Reviewer 1 cites.

For example, Barfield, Hold, and Gomulkiewicz (2011) develop a general model of eco-evolutionary dynamics in a stage-structured population. Their model is highly general, defining a multivariate quantitative trait with functions that define transition probabilities across different life stages; it makes some reasonable simplifying assumptions and nicely demonstrates that Lande's result of phenotypes evolving at the same rates across ages also applies to stages. This is an important and exciting theoretical result, but the authors also recognise the importance of putting it in the context of the fundamental equation of evolution: "In this section, we show how evolution in stage-structured populations can be related to a simple, universal law of evolution, Price's theorem (Price 1970), which provides a general statement about any evolutionary system -- including a stage-structured one". It is perhaps relevant to note that Price's (1970) motivation was solely to present a transparent "tool in qualitative evolutionary reasoning".

Lion, Sasaki, and Boots (2023) outline a theoretical framework linking quantitative genetics and adaptive dynamics, which reviews classical approaches to investigating eco-evolutionary dynamics and outlines key challenges for eco-evolutionary modelling. Rees and Ellner (2016) use an integral project model and compare it to an individual-based model (IBM), demonstrating how the former can be used to show why population structure changes are important for selection and generally recover results in the IBM. Coulson and Tuljapurkar (2008) investigate how selection on trait change can be partitioned into different components using an extension of the Price (1970) equation that accounts for age structure.

We don't just point out these particular examples because they are written by authors that Reviewer 1 cites. These examples are also highly general models (with the ones cited in our original manuscript being especially general examples), and typically dynamically sufficient in a way that leads to interesting new eco-evolutionary predictions. They build off of very general functions as a starting point, as in Lion (2018) and Patel et al. (2018), or expand upon more fundamental equations such as that of Price (1970). But these models all have a different goal from ours; to be useful, they make simplifying assumptions to link ecological and evolutionary dynamics.

Reviewer 1 is not correct in stating that these models retrieve the Price equation; the Price equation is not a predictive model but a mathematical identity that formally defines evolutionary change. Hence, it must always be possible, at least in theory, to retrieve the Price equation from any evolutionary model. These models don't retrieve the Price equation so much as the Price equation is necessarily embedded within them. The Price equation is not a simplified version of, e.g., Coulson and Tuljapurkar (2008), which includes a version of the Price equation that takes into account stage structure -- the latter is a specific case that is derived from the more abstract and exact former by making useful simplifying assumptions to generate novel predictions. The Price equation is also not just a generic mathematical function incorporating relevant evolutionary variables (as in, e.g., Lion and Patel examples mentioned above). Price (1970) is derived from first principles in evolutionary theory to define evolution in the same way that N_{t+1} = N_{t} + Births - Deaths defines ecological change.

Similarly, we attempt to define what is most fundamental to eco-evolutionary change. Our equation 1 does this formally, defining eco-evolutionary change in its most abstract and exact form as an interaction between fitness across different demographic components and individual characteristics. From eqn 1, the exact equations defining population and evolutionary change can be recovered. The simplicity of our equation is deceptive, and this is underscored by the fact that despite having never been discovered (to our knowledge), the equation can be used to derive the two most fundamental equations in ecology and evolution. As noted by Frank (1995, p. 381) in reference to Price's work, “[i]t is always difficult, in retrospect, to see the originality and insight of a simple idea”. While we certainly do not claim our work here to be field-changing, or as versatile as modelling tools from the abovementioned papers, we do believe that it is original and insightful in defining eco-evolutionary change.

In summary, are familiar with the literature pointed out by Reviewer 1, and we would even (quite happily) argue that this literature includes models that are highly general tools for modelling complex eco-evolutionary dynamics. The whole point of introducing the paper the way that we did was to illustrate that we are trying to ask a different question than the ones in those papers. We obviously need to try a different approach, and we have done this using extensive revisions in our introduction that address the points made above.

Reviewer 1 comment 2
==================================

The equation that is developed is not dynamically sufficient. This means that without modification it is fairly limited in describing what happens over an individual time step. This is no surprise, because the Price is simply a tautology that is also not dynamically sufficient. The authors may have derived an equation that is worthy of publication. They need to be honest on its limitations (also see point 1).

Author response to Reviewer 1 comment 2
==================================

We now state this more clearly in the text (Line XXX).

Reviewer 1 is of course correct that our eqn 1 is not dynamically sufficient. It couldn't possibly be, for exactly the reason that Reviewer 1 points out. The Price equation is a tool for unifying evolutionary models and providing conceptual clarity, which is necessarily achieved at the cost of any and all predictive power for a given system. Models need to make (simplifying, specific system) assumptions to be predictive (Levins 1966, Frank 2012), and a fundamental equation, by its nature, needs to be free of such assumptions and focus only on what is essential. The worthiness of our equation rests entirely on its power to unify fundamental equations of ecology and evolution and provide conceptual clarity, and its limitations are the same as any other fundamental equation.

Reviewer 1 comment 3
==================================

The authors massively overstate the importance of their equation, starting with the title of "Darwin's dream". There is no evidence I am aware of that Darwin ever thought much about equations. The authors then talk about scientific breakthroughs, attempting to place their equation with other great equations of science. Astonishing self-belief, and not warranted given how much has been done in this space already and the limitations of their formulae. The paper needs toning down significantly. The equation is mildly interesting, but it is not field-changing. This self belief may come from the authors having been working in an echo chamber. This is even hinted at in the acknowledgements and author contributions.

If the authors were to say they have found a way to write an equation that links the Price equation to the equation that describes change in population size from one time step to the next, that would be an honest and critical placement of their work in the broader literature. Explaining why it might be useful would help too.

Author response to Reviewer 1 comment 3
==================================

We have changed the title and reframed the introduction. Our goal was to explain the importance of conceptual unification in the sciences (which, in our experience, is often underappreciated in the life sciences) without overselling our present results. In our revised introduction, we have attempted to tone down our own contribution while also not losing our central point that this kind of theory is important in science (i.e., *useful*, but because it demonstrates logical consistency and provides conceptual clarity -- not just because you can use it in the same way as a predictive model).

In our manuscript, we argue that our equation is fundamental to ecology and evolution. What we mean by this is that the fundamental equations of ecology and evolution can be derived from our eqn 1. What this illustrates is a logical consistency for developing theory, which has previously been assumed but never actually derived. It does not imply that our equation is more *important* than any other equation, and this might be at least partly responsible for what Reviewer 1 interpreted as an inflated self-belief and need to tone the writing down. We do not believe our work to be field-changing, but we do believe that it is more than mildly interesting, and that it is necessary for reasons that are not well-appreciated.

What we have shown is a logical consistency between the fundamental equations of ecology and evolution (and, as far as we can tell, this is not disputed by either reviewer). This logical consistency has been assumed by theoreticians, and perhaps quite reasonably. But as a discipline, it is still important to do our due diligence and show that the theoretical foundations of ecology and evolution are solid. For example, within evolutionary biology, we can model evolutionary change using a population genetics framework or a quantitative genetics framework. In a population genetics framework, the average excess equation describes selection in terms of changes in allele frequencies, whereas in the quantitative genetics framework, the Breeder's equation describes selection in terms of changing quantitative traits (Queller 2017). These two equations are very important for modelling evolutionary change and predicting the consequences of selection. But an important role of theory is to demonstrate how (and why) these two equations are logically consistent, i.e., that they follow without contradiction from the same starting point that formally defines evolutionary change in any system. The Price equation and the framework that it establishes (Gardner 2020, Baravalle et al. 2025) is that starting point; it defines what evolution is in the broadest sense, meaning that any specific model of evolutionary change must be possible to derive from it. If a model of evolution cannot be put into the framework of the Price equation -- at least in principle -- then it is fundamentally flawed.

Hence, while the Price equation has very little direct relevance to the work of most evolutionary biologists, it is fundamental in the sense that it binds together all of evolutionary theory by defining what evolutionary change is (and is not). Our equation 1 fulfils the same criteria: from a deceptively simple starting point, eqn 1 can be used to derive the Price equation itself, but also the fundamental equation of population ecology. It therefore defines what eco-evolutionary change is (and is not), mathematically, at its most fundamental. Verbally, it is the interaction between absolute fitness and some characteristic of an individual.

The fact that the theoretical foundations of ecology and evolution are logically consistent is, of course, hardly a surprise. Had this not been the case, theoretical biologists would surely have noticed by now. But demonstrating this fact from first principles is, we believe, a worthy goal in itself, and something that we have considered exciting work.

Reviewer 2 comment 1  
======================

At the broadest level, I've always understood the Price Equation as explicitly including the birth-death process that would lead to the population growth rate equation from ecology. In the abstract sense, the Price Equation does not require ancestors and descendants to be different, and thus the relation between them necessarily captures births (when they are different) and deaths (when the ancestor is not among the descendants). This relationship allows mean fitness to capture the population growth rate, as first shown by Fisher (1930) and later by folks like Lande (1982) and Rice (2004). Day & Bonduriansky (2011) note that the total number of individuals in each time interval is the sum of all surviving parents and offspring; hence, they are Nt(births – deaths), which is Duthie & Luque’s equation 3. Now, a case could be made that stating this relationship in an obvious and explicit way is useful, but it seems to me that the original formulation of the Price Equation already subsumes population growth dynamics, so erecting an even broader equation isn’t necessary.

Author response to Reviewer 2 comment 1 
=====================

In some sense, it is true that the Price equation subsumes population change; if it did not, then we would not be able to produce a unifying equation from which both the Price equation and the general Nt(birhts - deaths) equation could be derived. However, in the traditional derivation of the Price equation, births, deaths, and population growth are not at all explicit. Instead, the number of individuals (or groups or genes -- whatever the entity is) is simply defined as the sum across n_{i} at t and n'_{i} at t + 1, and frequencies are then immediately defined by n_{i}/n and n'_{i}/n' (Price 1970; Frank 1997; 2012; Gardner 2008). Numbers of entities (n and n') are only mentioned so that the frequencies of entities (q_{i} and q'_{i}) can then be defined, and these frequencies are then used to work out entity fitness. To our knowledge, the Price equation has never been derived from births and deaths or with reference to population growth. Lande (1982) did introduce a demographic model for population demography and dN/dt = rN does appear (eqn 10), but this is in the context of a modelling assumption; the Price equation is then used to calculate a selection gradient. Similar to previously referenced eco-evolutionary models (see our response to Reviewer 1 comments), there is integration of ecology and evolution here, but not a demonstration of the key link from first principles. Similarly, Day and Bonduriansky (2011) define a value W that is the sum of offspring and survival, and this is used within a version of the Price equation to get mean fitness, but a formal link to population growth isn't made, with the goal instead being to predict evolving traits.

Overall, population change appears in models where the Price equation also appears, but what has always been missing is a derivation of both from the same starting point, from first principles of what actually happens in a population. Logically, this has to be possible because the same mechanisms that underlie population change (birth and death) also underlie evolutionary change. But up until now, to the best of our knowledge, the link between population and evolutionary change has always been carefully considered but ultimately ad hoc (as in the abovementioned references and, to our knowledge, all eco-evolutionary models). Given that we are looking at the same mechanisms of population change, we really should be able to demonstrate that both population change and evolutionary change can 'pop out' as intrinsically and necessarily linked from the same starting point.

Figuring this starting point out was not easy; it needed to be done deductively, which inevitably led to many dead ends. Initially, we reasoned that because the Price equation is sufficiently abstract to describe trait change in any population (including groups and higher moments of traits), its recursive properties, and how useful it is for making formal links across disciplines such as physics and information theory (Frank 2015; 2016), we should be able to use it directly to recover population density change. This ended up being a dead end for reasons we will not dwell on (briefly, conservation of total probability -- that entity frequencies need to sum to 1 -- precludes the Price equation from being used to directly to recover absolute population change). We had different issues when starting from population change and trying to recover the Price equation. There are a lot of ways to make coherent models using these avenues of inquiry, but none that recover universal and exact descriptions of population and evolutionary change. The broader equation, our equation 1, is indeed necessary to derive exact and universal eco-evolutionary change from first principles; it thereby serves as a formal definition for eco-evolutionary change in any system. It perhaps appears simple and obvious in hindsight, as if it could be inferred quickly from existing models by anyone giving it a modicum of thought. But we believe this appearance is highly deceptive; if this formal link were trivial to discover, then someone would have surely pointed it out already. After much investigation, we are confident that no one has, and while there are fundamental equations defining ecological and evolutionary change, separately, our work here is unique in establishing a fundamental equation that defines eco-evolutionary change.

In our revision, we try to clarify the need for our equation more clearly.

Reviewer 2 comment 2
======================

This leads me to my second point. The broader equation (1) is defined such that it introduces an abstract quantity (Omega), whose meaning depends on further definitions of variables. In its original formulation (1), it is meaningless. For example, if N = 3, beta = 1, delta = 1, z1 = 1, z2 = 2, and z3 = 3, and Deltaz = 0.5, then the first term in equation 1 reduces to 1, and the summation becomes a sum of z: 1.5 + 2.5 + 3.5 = 7.5, thus, Omega is the sum of z. But if the first term doesn’t reduce to 1, Omega becomes incomprehensible. Let’s say beta_1 = 2, with the other betas remaining as 1. Now,

Omega = (2-1 + 1)(1 + 0.5) + (1-1 + 1)(2 + 0.5) + (1-1 + 1)(3 + 0.5) = 10.5

### 2025-05-29 foundations_of_ecoevo.Rmd
Initialise response to reviewer comments, with current manuscript

---
title: "Darwin's Dream: Unifying ecological and evolutionary change"
author: |
  | A. Bradley Duthie$^{1,a,*}$ and Victor J. Luque$^{2,a}$
date: '[1] Department of Biological and Environmental Sciences, University of Stirling, Stirling, Scotland [2] Department of Philosophy, University of Valencia, Valencia, Spain [*] Corresponding author: alexander.duthie@stir.ac.uk [a] Equal contribution'
output:
  pdf_document:
      keep_tex: yes
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 2
bibliography: refs.bib
link-citations: yes
linkcolor: blue
biblio-style: apalike
---

@@

**Biological evolution is realised through the same mechanisms of birth and death that underlie change in population density.**
**The deep interdependence between ecology and evolution is well-established, but much theory in each discipline has been developed in isolation.**
**While recent work has accomplished eco-evolutionary synthesis, a gap remains between the logical foundations of ecology and evolution.**
**We bridge this gap with a new equation that defines a summed value for a characteristic across individuals in a population, from which the fundamental equations of population ecology and evolutionary biology (the Price equation) are derived.**
**We thereby unify the fundamental equations of population ecology and biological evolution under a general framework.**
**Our unification further demonstrates the equivalence between mean population growth rate and evolutionary fitness, shows how ecological and evolutionary change are reflected in the first and second statistical moments of fitness, respectively, and links this change to ecosystem function.**
**Finally, we outline how our proposed framework can be used to unify social evolution and density-dependent population growth.**

@@

@@

Biologists have long known that the ecology of population size is inextricably linked with the evolution of population traits. Nevertheless, the foundational theory underlying ecological and evolutionary change have been historically separated, with recent attempts to bridge this theoretical divide using eco-evolutionary modelling. We discover a new equation from first principles that provides a universal and complete definition for eco-evolutionary change. This equation thereby offers a unified framework for ecology and evolution, which also recovers the foundational theory of ecology and evolution as developed in isolation.

@@

Theoretical unification of phenomena is a powerful tool for scientific advancement. 
Such unification has been a major goal in scientific research throughout history [@Kitcher1993], and its value is perhaps most evident in reconciling unconnected models and revealing new and unexpected empirical predictions. 
In theoretical physics, Newton's unification of celestial and terrestrial motion, and Maxwell's unification of electricity and magnetism, are among the most important advances in scientific history. 
Likewise, in the life sciences, the modern synthesis of natural selection and Mendelian inheritance revolutionised how mechanisms underlying evolution and genetics are understood and applied [@Gayon2019].
The analogy to physics is imperfect because evolution encompasses non-biological systems with non-genetic inheritance [e.g., @Price1995; @Shennan2011; @MacCallum2012; @Lewens2015], but to illustrate our point, we restrict our focus to biological evolution.
Following the modern synthesis, patterns in evolutionary biology and genetics could be interpreted through a common framework.

@@

In contrast, theory underlying ecology and evolution has not been so tightly intertwined.
Until recently, population ecology and evolution in particular have taken almost parallel paths for developing theory.
Although many revelatory models exist to connect ecological and evolutionary phenomena [e.g., @Lion2018; @Patel2018; @Govaert2019], the foundational assumptions and equations of ecology and evolution in these models remain mostly separated conceptually. 
But like the mechanisms driving evolutionary and genetic change, those that drive ecological and evolutionary change are well-established with predictable joint consequences.
In evolution, natural selection, drift, and gene flow are all instantiated through a combination of different ecological factors [@Connor2004].
Most notably, change in population density and phenotype  are both caused by individual birth and death [@Turchin2001].
This should allow biologists to place ecological and evolutionary change on a shared theoretical foundation.

@@

To fully unify biological evolution and population ecology, we must reconcile the general equation for population change (Box 1) with the Price equation (Box 2).
Previous theory has produced mathematical synthesis between ecological and evolutionary equations [e.g., @Page2002; @Collins2009], but not established a unified conceptual framework for joint ecological and evolutionary change.
For example, @Page2002 demonstrate a mathematical equivalence between the Price equation and a general equation for community dynamics. 
In both equations, the relative frequencies of different ecological (species) or evolutionary (phenotype) variants map to their respective ecological (population growth) or evolutionary (fitness) change in a way that is mathematically equivalent.
@Collins2009 use the Price equation to build a general quantitative framework that partitions phenotypic change into ecological, evolutionary, and physiological components.
The average change in a community-wide trait (e.g., carbon uptake) is exactly and generally partitioned into change attributable to changes in species composition, expected evolution of lineages within species, and expected change in lineage physiology across all species.
Both @Page2002 and @Collins2009 use relative species abundances, meaning that species frequencies must sum to one.
This is useful for demonstrating mathematical equivalence [@Page2002] or expected change in a community-wide trait [@Collins2009], but a different approach is needed to recover absolute change in population size and thereby recover a joint foundation for ecology and evolution.

The Price equation is critical for partitioning different components of biological change [@Price1970; @Frank1997; @Gardner2008; @Luque2016; @Lehtonen2018; @Queller2017].
It has also been highly useful for integrating evolutionary theory across disciplines [@Fox2006; @Brantingham2007; @MacCallum2012; @Frank2015; @Godsoe2021; @Ulrich2024].
These properties would seem to make it an intuitive starting point for a logical foundation of ecology and evolution, perhaps through some kind of mathematical equivalence [e.g., @Page2002] or addition of terms [e.g., @Collins2009], or through the use of its recursive structure [e.g., @Kerr2009; @Frank2012].
But despite its flexibility, the Price equation still relies on relative frequencies, which must by definition sum to one [@Frank2015].
This is because the Price equation describes the average change in a population; the frequency of entities is scaled thereby conserving total probability [@Frank2015; @Frank2016].
But to recover the fundamental principle of exponential population growth [@Turchin2001], this scaling must be avoided in a fundamental equation of ecology and evolution.

@@

@@

We therefore recover the general equation for population ecology (eqn 2) and the fundamental property of exponential growth in populations [eqn 3, @Turchin2001].

@@

The Price equation provides a complete and exact description of evolution in any closed evolving system [@Price1970; @Frank2012].
It is derived by rearranging the mathematical notation defining changes in the frequencies and characteristics of any type of entity [e.g., individuals, alleles, @Price1970; @Gardner2008; @Luque2016].
This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation [e.g., group, individual, @Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
Because of its abstract nature and lack of any system-specific assumptions, the Price equation makes no predictions about what will happen in any particular system [@Gardner2020].
Its role is not to predict, but to formally and completely define and separate components of evolutionary change. 
The same is true of the general equation for population change (eqns 2 and 3), at least as we have used it here where it serves to define what population growth means in ecology.
This equation formally and completely describes population change in terms of births and deaths.
In eqn 1, we therefore have a fundamental equation from which we can derive complete ecological and evolutionary change in any closed biological population.
We anticipate that this will be useful for eco-evolutionary theory in the same way that the Price equation is useful for evolutionary theory: facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements.

Our unification demonstrates the equivalence between the finite rate of increase $\lambda$ [Box 1, @Gotelli2001] and population mean evolutionary fitness $\bar{w}$ (Box 2).
The population growth equation $N_{t+1} = N_{t}\lambda$ could therefore be rewritten as $N_{t+1} = N_{t}\bar{w}$.
We do not claim to be the first to notice this equivalence between population growth and mean population fitness.
Indeed, the relationship between population growth rate and evolutionary fitness has been proposed and applied many times before [e.g., @Fisher1930; @Charlesworth1980; @Lande1982; @Roff2008; @Lion2018].
For example, @Lande1982 explicitly concludes mean absolute fitness per unit time is $\bar{w} = e^{r}$, where $r = ln(\lambda)$.
We show this from first principles and clarify the relationship between fitness and population growth.
Over an arbitrary length of time, fitness is properly defined as $w_{i} = \beta_{i} - \delta_{i} + 1$.
Over an individual's lifetime (which, by definition, includes death), fitness is therefore $\beta_{i}$.
Interestingly, the rate of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively.
Population growth rate reflects mean fitness $\bar{w}$, while the rate of evolutionary change reflects the variance in fitness $\mathrm{Var}(w)$ [i.e., Fisher's fundamental theorem, @Frank1997; @Rice2004; @Queller2017].

>>>>

---
title: "Darwin's Dream: Unifying ecological and evolutionary change"
author: |
  | A. Bradley Duthie$^{1,a,*}$ and Victor J. Luque$^{2,a}$
date: '[1] Department of Biological and Environmental Sciences, University of Stirling, Stirling, Scotland [2] Department of Philosophy, University of Valencia, Valencia, Spain [*] Corresponding author: alexander.duthie@stir.ac.uk [a] Equal contribution'
output:
  word_document: default
  pdf_document:
      keep_tex: yes
  html_document:
    df_print: paged
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 2
bibliography: refs.bib
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
biblio-style: apalike
---

@@

Biological evolution is realised through the same mechanisms of birth and death that underlie change in population density. 
The deep interdependence between ecology and evolution is well-established, but much theory in each discipline has been developed in isolation. 
While recent work has accomplished eco-evolutionary synthesis, a gap remains between the logical foundations of ecology and evolution. 
We bridge this gap with a new equation that defines a summed value for a characteristic across individuals in a population, from which the fundamental equations of population ecology and evolutionary biology (the Price equation) are derived. 
We thereby unify the fundamental equations of population ecology and biological evolution under a general framework. 
Our unification further demonstrates the equivalence between mean population growth rate and evolutionary fitness, shows how ecological and evolutionary change are reflected in the first and second statistical moments of fitness, respectively, and links this change to ecosystem function. 
Finally, we outline how our proposed framework can be used to unify social evolution and density-dependent population growth.

@@

<!---

@@

Biologists have long known that the ecology of population size is inextricably linked with the evolution of population traits. Nevertheless, foundational theories underlying ecological and evolutionary change have been historically separated, with recent attempts to bridge the theoretical divide using eco-evolutionary modelling. We discover a new equation from first principles that provides a universal and complete definition for eco-evolutionary change. This equation thereby offers a unified framework for ecology and evolution, which also recovers the foundational theory of ecology and evolution as developed in isolation.

--->

@@

Theoretical unification of phenomena is a powerful tool for scientific advancement. 
Such unification has been a major goal in scientific research throughout history [@Kitcher1993], and its value is perhaps most evident in reconciling unconnected models and revealing new and unexpected empirical predictions. 
In theoretical physics, Newton's unification of celestial and terrestrial motion, and Maxwell's unification of electricity and magnetism, are among the most important advances in scientific history. 
Likewise, in the life sciences, the modern synthesis of natural selection and Mendelian inheritance revolutionised how mechanisms underlying evolution and genetics are understood and applied [@Gayon2019].
The analogy to physics is imperfect because evolution encompasses non-biological systems with non-genetic inheritance [@Price1995; @Shennan2011; @MacCallum2012; @Lewens2015], but to illustrate our point, we restrict our focus to biological evolution.
Following the modern synthesis, patterns in evolutionary biology and genetics could be interpreted through a common framework.

@@

In contrast, theory underlying ecology and evolution has not been so tightly intertwined.
Until recently, population ecology and evolution in particular have taken almost parallel paths for developing theory.
Although many revelatory models exist to connect ecological and evolutionary phenomena [@Lion2018; @Patel2018; @Govaert2019], the foundational assumptions and equations of ecology and evolution in these models remain mostly separated conceptually. 
But like the mechanisms driving evolutionary and genetic change, those that drive ecological and evolutionary change are well-established with predictable joint consequences.
In evolution, natural selection, drift, and gene flow are all instantiated through a combination of different ecological factors [@Connor2004].
Most notably, change in population density and phenotype  are both caused by individual birth and death [@Turchin2001].
This should allow biologists to place ecological and evolutionary change on a shared theoretical foundation.

@@

To fully unify biological evolution and population ecology, we must reconcile the general equation for population change (Box 1) with the Price equation (Box 2).
Previous theory has produced mathematical synthesis between ecological and evolutionary equations [@Page2002; @Collins2009], but not established a unified conceptual framework for joint ecological and evolutionary change.
For example, @Page2002 demonstrate a mathematical equivalence between the Price equation and a general equation for community dynamics. 
In both equations, the relative frequencies of different ecological (species) or evolutionary (phenotype) variants map to their respective ecological (population growth) or evolutionary (fitness) change in a way that is mathematically equivalent.
@Collins2009 use the Price equation to build a general quantitative framework that partitions phenotypic change into ecological, evolutionary, and physiological components.
The average change in a community-wide trait (e.g., carbon uptake) is exactly and generally partitioned into change attributable to changes in species composition, expected evolution of lineages within species, and expected change in lineage physiology across all species.
Both @Page2002 and @Collins2009 use relative species abundances, meaning that species frequencies must sum to one.
This is useful for demonstrating mathematical equivalence [@Page2002] or expected change in a community-wide trait [@Collins2009], but a different approach is needed to recover absolute change in population size and thereby recover a joint foundation for ecology and evolution.

The Price equation is critical for partitioning different components of biological change [@Price1970; @Frank1997; @Gardner2008; @Luque2016; @Lehtonen2018; @Queller2017].
It has also been highly useful for integrating evolutionary theory across disciplines [@Fox2006; @Brantingham2007; @MacCallum2012; @Frank2015; @Godsoe2021; @Ulrich2024].
These properties would seem to make it an intuitive starting point for a logical foundation of ecology and evolution, perhaps through some kind of mathematical equivalence [@Page2002] or addition of terms [@Collins2009], or through the use of its recursive structure [@Kerr2009; @Frank2012].
But despite its flexibility, the Price equation still relies on relative frequencies, which must by definition sum to one [@Frank2015].
This is because the Price equation describes the average change in a population; the frequency of entities is scaled thereby conserving total probability [@Frank2015; @Frank2016].
But to recover the fundamental principle of exponential population growth [@Turchin2001], this scaling must be avoided in a fundamental equation of ecology and evolution.

@@

<!---

19 MAY: Point out here that by itself, Omega has the same terms as z.

- Verbally, our equation shows interaction between fitness and an individual's characteristic, summed
- Because it is summing across individuals, some characteristics can be interpreted as measure of ecosystem function
- Ecosystem function will make the assumption of additivity of z_i across all N
- We will see that mean fitness is population ecology
- Variance is evolutionary change
- Summation is ecosystem function

Eqn 1 captures an invariant relationship between individual fitness and an individual's characteristic

Also note that directly tying fitness to demography

--->

@@

We therefore recover the general equation for population ecology (eqn 2) and the fundamental property of exponential growth in populations [@Turchin2001] (eqn 3).

@@

The Price equation provides a complete and exact description of evolution in any closed evolving system [@Price1970; @Frank2012].
It is derived by rearranging the mathematical notation defining changes in the frequencies and characteristics of any type of entity [@Price1970; @Gardner2008; @Luque2016] (e.g., individuals, alleles).
This derivation partitions total characteristic change into different components, making it possible to isolate evolutionary mechanisms (e.g., selection) and levels of biological organisation (e.g., group, individual) [@Frank1995; @Kerr2009; @Frank2012; @Luque2016; @Okasha2020].
Because of its abstract nature and lack of any system-specific assumptions, the Price equation makes no predictions about what will happen in any particular system [@Gardner2020].
Its role is not to predict, but to formally and completely define and separate components of evolutionary change. 
The same is true of the general equation for population change (eqns 2 and 3), at least as we have used it here where it serves to define what population growth means in ecology.
This equation formally and completely describes population change in terms of births and deaths.
In eqn 1, we therefore have a fundamental equation from which we can derive complete ecological and evolutionary change in any closed biological population.
We anticipate that this will be useful for eco-evolutionary theory in the same way that the Price equation is useful for evolutionary theory: facilitating specific model development and identifying new conceptual insights, unresolved errors, and sources of model disagreements.

Our unification demonstrates the equivalence between the finite rate of increase $\lambda$ (Box 1) [@Gotelli2001] and population mean evolutionary fitness $\bar{w}$ (Box 2).
The population growth equation $N_{t+1} = N_{t}\lambda$ could therefore be rewritten as $N_{t+1} = N_{t}\bar{w}$.
We do not claim to be the first to notice this equivalence between population growth and mean population fitness.
Indeed, the relationship between population growth rate and evolutionary fitness has been proposed and applied many times before [@Fisher1930; @Charlesworth1980; @Lande1982; @Roff2008; @Lion2018].
For example, @Lande1982 explicitly concludes mean absolute fitness per unit time is $\bar{w} = e^{r}$, where $r = ln(\lambda)$.
We show this from first principles and clarify the relationship between fitness and population growth.
Over an arbitrary length of time, fitness is properly defined as $w_{i} = \beta_{i} - \delta_{i} + 1$.
Over an individual's lifetime (which, by definition, includes death), fitness is therefore $\beta_{i}$.
Interestingly, the rate of change in ecology and evolution are reflected in the first and second statistical moments of fitness, respectively.
Population growth rate reflects mean fitness $\bar{w}$, while the rate of evolutionary change reflects the variance in fitness $\mathrm{Var}(w)$ (i.e., Fisher's fundamental theorem) [@Frank1997; @Rice2004; @Queller2017].
